Circularity in Population-Based FA Analysis

### 2011-07-11 ANTsTBSS/paper.tex
ENH: Minor edits.
Added refernce.

\documentclass[preprint,authoryear,review,12pt]{elsarticle}
%\documentclass[final,5p,times,twocolumn]{elsarticle}

@@

Although various processing choices mitigate these effects, such as Gaussian smoothing, nonparametric testing, and the use of more robust similarity metrics, such choices are often made on an ad hoc basis.  
In contrast, in this work we strongly advocate the general application
%A more principled approach is the use of an anatomical, i.e. T1-weighted, unbiased shape/intensity 
%template constructed from the population images.  DTI-derived scalar images can then be mapped 
%to their corresponding T1-weighted image in a highly constrained manner and subsequently 
%mapped to the template via nonrigid registration of the T1-weighted image.  This minimizes 
%possible complications associated with direct FA-to-FA image registration.
%We recognize that using the anatomical images in the alignment process has been previously proposed (in fact, such a suggestion occurs in the {\it Future directions} section of the original TBSS paper).  
of a high performance \citep{Klein2009}, nonrigid registration algorithm \citep{Avants2011}, which forms the underlying machinery for an optimal template construction algorithm \citep{Avants2010}, in facilitating alignment between DTI-derived scalar images for cross-population studies.  The key point, here, is that the FA images are mapped to the common reference space via intra-subject DT-to-T1 mapping followed by T1-to-T1 alignment.  The T1-to-T1 step is (statistically) independent of the FA values which are the ultimate statistical target of hypothesis testing in FA population studies.  Below, we show that this same pattern of effects is reproduced in real data and detail our proposed solution.

@@

\begin{table}
\begin{center}
  \caption{Overlap measures of thresholded mean FA and MNI152 white matter.}
  \label{table:lom}
  \begin{tabular*}{130mm}{@{\extracolsep{\fill}} cccccc}
  \hline
%  {} & \multicolumn{2}{c}{NKI} & \multicolumn{2}{c}{NYU} \\
%  \cline{2-5}
  {Norm.} & Target & Jaccard & Dice  & False $-$ & False $+$\\
  \hline
  \hline
  antsT1 & 0.917 & 0.621 & 0.767 & 0.082 & 0.341 \\
  antsCC & 0.780 & 0.570 & 0.727 & 0.220 & 0.320 \\
  antsMI & 0.776 & 0.569 & 0.725 & 0.223 & 0.320 \\
  antsDemons & 0.780 & 0.566 & 0.723 & 0.220 & 0.326 \\
  antsSSD & 0.783 & 0.562 & 0.720 &  0.217 & 0.334 \\
  tbss & 0.716 & 0.550 & 0.710 & 0.283 & 0.295 \\
  \hline
  \end{tabular*}
\end{center}  
\end{table}

@@

Caveats with our analysis include a lack of ground truth in the
clinical component of the study.   Additionally, we are not addressing
the biological plausibility or our results nor do we
intend to (in this paper) enter into discussion on the mechanisms of
TBI, gender difference and DTI that may lead to the detected results.
Rather, we focus on the technical issue of circularity caused by
normalizing the image feature of interest with functions that maximize
the statistic used in hypothesis testing.  We also cannot argue that our analysis
invalidates previous studies in which circularity might be an issue.  
Finally, we do not address the very promising work concerning
whole tensor DTI normalization \citep{Zhang2007,Hecke2007} and DTI template-based
strategies \citep{Mori2009,Hecke2011}.  Although investigation is beyond the scope
of this work, these whole-tensor approaches might provide the necessary
anatomical information to overcome the deficiencies addressed in this
work.

>>>>

%\documentclass[preprint,authoryear,review,12pt]{elsarticle}
\documentclass[final,5p,times,twocolumn]{elsarticle}

@@

Although various processing choices mitigate these effects, such as Gaussian smoothing, nonparametric testing, and the use of more robust similarity metrics, such choices are often made on an ad hoc basis.  
In contrast, in this work we strongly advocate the general application
%A more principled approach is the use of an anatomical, i.e. T1-weighted, unbiased shape/intensity 
%template constructed from the population images.  DTI-derived scalar images can then be mapped 
%to their corresponding T1-weighted image in a highly constrained manner and subsequently 
%mapped to the template via nonrigid registration of the T1-weighted image.  This minimizes 
%possible complications associated with direct FA-to-FA image registration.
%We recognize that using the anatomical images in the alignment process has been previously proposed (in fact, such a suggestion occurs in the {\it Future directions} section of the original TBSS paper).  
of a high performance \citep{Klein2009}, nonrigid registration algorithm \citep{Avants2011}, which forms the underlying machinery for an optimal template construction algorithm \citep{Avants2010}, in facilitating alignment between DTI-derived scalar images for cross-population studies.  The key point is that the FA images are mapped to the common reference space via intra-subject DT-to-T1 mapping followed by T1-to-T1 alignment.  The T1-to-T1 step is (statistically) independent of the FA values which are the ultimate statistical target of hypothesis testing in FA population studies.  Below, we show that this same pattern of effects is reproduced in real data and detail our proposed solution.

@@

\begin{table}
\begin{center}
  \caption{Overlap measures of thresholded mean FA and MNI152 white matter.}
  \label{table:lom}
  \begin{tabular*}{85mm}{@{\extracolsep{\fill}} cccccc}
  \hline
%  {} & \multicolumn{2}{c}{NKI} & \multicolumn{2}{c}{NYU} \\
%  \cline{2-5}
  {Norm.} & Target & Jaccard & Dice  & False $-$ & False $+$\\
  \hline
  \hline
  antsT1 & 0.917 & 0.621 & 0.767 & 0.082 & 0.341 \\
  antsCC & 0.780 & 0.570 & 0.727 & 0.220 & 0.320 \\
  antsMI & 0.776 & 0.569 & 0.725 & 0.223 & 0.320 \\
  antsDemons & 0.780 & 0.566 & 0.723 & 0.220 & 0.326 \\
  antsSSD & 0.783 & 0.562 & 0.720 &  0.217 & 0.334 \\
  tbss & 0.716 & 0.550 & 0.710 & 0.283 & 0.295 \\
  \hline
  \end{tabular*}
\end{center}  
\end{table}

@@

Caveats with our analysis include a lack of ground truth in the
clinical component of the study.   Additionally, we are not addressing
the biological plausibility or our results nor do we
intend to (in this paper) enter into discussion on the mechanisms of
TBI, gender difference and DTI that may lead to the detected results.
Rather, we focus on the technical issue of circularity caused by
normalizing the image feature of interest with functions that maximize
the statistic used in hypothesis testing.  We also cannot argue that our analysis
invalidates previous studies in which circularity might be an issue.  
Finally, we do not address the very promising work concerning
whole tensor DTI normalization \citep{Zhang2007,Hecke2007}, DTI template-based
strategies \citep{Mori2009,Hecke2011}, {\color{red}{or other related methods \cite{jbabdi2010}}}.  Although investigation is beyond the scope
of this work, these whole-tensor approaches might provide the necessary
anatomical information to overcome the deficiencies addressed in this
work.
